% !TeX root = ../QFT_main.tex
%=========================================================
%=========================================================
\chapter{Non-perturbative aspects of QFT}
%========================================================
%=========================================================



%--------------------------------------------------------
%=========================================================
\section{Kallén-Lehmann spectral representation}
%=========================================================
%------------------------------------------------------------



\begin{mytheorem}[The states of a theory (Matheus 2.1)]
In order to discuss the Kallén-Lehmann representation, we need to understand the states of a quantum field theory.
We can have essentially three types of states, the first two being essentially the same,
\begin{itemize}
    \item \textbf{Single-particle states.} These are states that correspond to a single excitation of the field, typically associated with a pole in the propagator of the theory.
    These have a well-defined mass and momentum, and are reasonably stable to be identified as particles.
    \item \textbf{Resonances.} These are states that correspond to a single excitation of the field, but are not stable enough to be identified as particles, and thus correspond to a resonance in the propagator of the theory.
    \item \textbf{Multi-particle states.} These correspond to multiple excitations of the field, i.e. a collection of single-particle states, typically associated with a branch cut in the propagator of the theory.
    Such a state have a well-defined total momentum, but not a well-defined mass, since the internal energy of the state depends on the relative momenta of the single-particle states that compose it.
    \item \textbf{Bound states.} These are states that correspond to a bound state of multiple particles, again typically associated with a pole in the propagator of the theory.
    These states do have a well-defined momentum and mass, which essentially corresponds to the sum of the masses of the constituent particles minus the binding energy.
\end{itemize}

We can label such states by $\lambda_p$ where $p$ is the total momentum of the state, and $\lambda$ is a label that distinguishes different states with the same momentum so that
\begin{align}
    |\lambda_p\rangle = U(p)|\lambda_0\rangle,
\end{align}
where $U(p)$ is the unitary operator implementing a boost to momentum $p$.
A resolution of the identity can then be written with Lorentz-invariant measure as
\begin{align}
    \mathbb{1} = \sum_\lambda \int \frac{d^3p}{(2\pi)^3}\frac{1}{2E_p}|\lambda_p\rangle\langle\lambda_p|\,.
\end{align}
\end{mytheorem}



\begin{mytheorem}[Kallén-Lehmann spectral representation (Matheus 2.2)]
Consider the 2-point Green function of a real scalar field $\phi$ in vacuum
\begin{align}
    G(x,y) = \langle 0|T\{\phi(x)\phi(y)\}|0\rangle\,.
\end{align}
Inserting a resolution of the identity in terms of the states of the theory, we can write...


\end{mytheorem}


%--------------------------------------------------------
%=========================================================
\section{Schwinger-Dyson equations}
%=========================================================
%------------------------------------------------------------


\begin{mytheorem}[Dyson--Schwinger equations for correlators \& equivalence of quantization schemes]
A quantum theory is completely specified by its correlators\footnote{In fact, it is even sufficient to only know time-ordered ones.} (expectation values of products of time-evolved field operators).
Correlators are uniquely determined by the Dyson--Schwinger equations, which are an infinite set of coupled differential equations relating all $n$-point functions to each other for all $n\in\mathbb{N}$.
If two theories have the same Dyson--Schwinger equations for all correlators, they are thus completely equivalent.
In particular this proves the equivalence between canonical quantization and path integral quantization, even in curved spacetime, at least as long as the former approach is applicable.
\end{mytheorem}