% !TeX root = ../QFT_main.tex
\documentclass[../QFT_main.tex]{subfiles}
\begin{document}
%=========================================================
%=========================================================
\chapter{Generating Functionals \& Path Integrals}\label{ch:path_integral_formulation}
%========================================================
%=========================================================



%--------------------------------------------------------
%=========================================================
\section{Path integral formulation in ordinary quantum mechanics}
%=========================================================
%------------------------------------------------------------

\begin{mytheorem}[Gaussian bosonic integrals]
For $\Re\alpha>0$, the elementary Gaussian integral and its square read as follows.
{\small\begin{align}\begin{aligned}
    I(\alpha) &:= \int_{\R}dx\,e^{-\alpha x^2} = \sqrt{\tfrac{\pi}{\alpha}}, \qquad I(1)^2 = \int_{\R^2}dx\,dy\,e^{-(x^2+y^2)} = \int_0^{2\pi}d\varphi\int_0^\infty dr\,r e^{-r^2} = \pi.
\end{aligned}\end{align}}
For a real symmetric positive-definite matrix $A\in\R^{n\times n}$, the multivariate generalization is the following.
{\small\begin{align}\begin{aligned}
    \int_{\R^n}d^nx\,e^{-\tfrac12 x^{\mathsf T}Ax} &= (2\pi)^{n/2}(\det A)^{-1/2}.
\end{aligned}\end{align}}
Adding a source $b$ amounts to shifting to the stationary point $x_c=-A^{-1}b$ and completing the square.
The shift and the resulting integral are summarized by the following identities.
{\small\begin{align}\begin{aligned}
    S(x) &:= \tfrac12 x^{\mathsf T}Ax+b^{\mathsf T}x = \tfrac12 (x-x_c)^{\mathsf T}A(x-x_c)-\tfrac12 b^{\mathsf T}A^{-1}b, \qquad \int_{\R^n}d^nx\,e^{-S(x)} = (2\pi)^{n/2}(\det A)^{-1/2}e^{\tfrac12 b^{\mathsf T}A^{-1}b}.
\end{aligned}\end{align}}
Interpreting $S$ as an action, $x_c$ is its classical stationary solution, since $\nabla S(x_c)=Ax_c+b=0$.
\end{mytheorem}



\begin{mytheorem}[Path integral formulation of quantum mechanics \`a la Feynman]
The transition amplitude between position eigenstates is the following matrix element.
{\small\begin{align}\begin{aligned}
    F(q_f,t_f;q_i,t_i) &:= {}_H\langle q_f,t_f|q_i,t_i\rangle_H = \langle q_f|e^{-\tfrac{i}{\hbar}\hat H(t_f-t_i)}|q_i\rangle_S.
\end{aligned}\end{align}}
Split $[t_i,t_f]$ into $N+1$ intervals of length $\epsilon=\tfrac{t_f-t_i}{N+1}$ and set $t_k=t_i+k\epsilon$.
\begin{center}
\begin{tikzpicture}[x=1.25cm,y=0.55cm,>=Stealth,font=\scriptsize]
    \draw[->] (-0.25,0) -- (5.45,0) node[right] {$t$};
    \draw[->] (0,-0.25) -- (0,3.1) node[above] {$q$};
    \foreach \x in {0,1,2,3,4,5} \draw[gray!55,densely dashed] (\x,0) -- (\x,2.8);
    \draw[blue!70!black,thick] (0,0.65) -- (1,2.25) -- (2,1.35) -- (3,2.6) -- (4,0.95) -- (5,1.85);
    \foreach \x/\y in {0/0.65,1/2.25,2/1.35,3/2.6,4/0.95,5/1.85} \fill[blue!70!black] (\x,\y) circle (1.5pt);
    \node[below] at (0,0) {$t_0=t_i$};
    \node[below] at (1,0) {$t_1$};
    \node[below] at (2,0) {$\cdots$};
    \node[below] at (3,0) {$t_k$};
    \node[below] at (4,0) {$\cdots$};
    \node[below] at (5,0) {$t_{N+1}=t_f$};
    \node[left] at (0,0.65) {$q_i$};
    \node[right] at (5,1.85) {$q_f$};
\end{tikzpicture}
\end{center}
Inserting position-space completeness relations gives the following skeleton.
{\small\begin{align}\begin{aligned}
    \mathbb{I} &= \int_{\R}dq_k\,|q_k,t_k\rangle_H{}_H\langle q_k,t_k|, \\
    F(q_f,t_f;q_i,t_i) &= \lim_{N\to\infty}\int\prod_{k=1}^{N}dq_k\prod_{k=0}^{N}{}_H\langle q_{k+1},t_{k+1}|q_k,t_k\rangle_H, \qquad q_0=q_i,\quad q_{N+1}=q_f.
\end{aligned}\end{align}}
Inserting $\mathbb{I}=\int_{\R}\tfrac{dp}{2\pi\hbar}|p\rangle\langle p|$ in each short-time factor and ordering every $\hat p$ to the left of every $\hat q$ gives the phase-space representation below.
{\small\begin{align}\begin{aligned}
    F(q_f,t_f;q_i,t_i) &= \lim_{N\to\infty}\int\prod_{k=1}^{N}dq_k\prod_{k=0}^{N}\tfrac{dp_{k+1}}{2\pi\hbar}\exp\Bigg\{\tfrac{i}{\hbar}\sum_{k=0}^{N}\Big[p_{k+1}(q_{k+1}-q_k)-\epsilon H(p_{k+1},q_k)\Big]\Bigg\} \\
    &= \int_{q(t_i)=q_i}^{q(t_f)=q_f}\mathcal{D}p\,\mathcal{D}q\,\exp\Bigg\{\tfrac{i}{\hbar}\int_{t_i}^{t_f}dt\,\big[p\dot q-H(p,q)\big]\Bigg\}.
\end{aligned}\end{align}}
The intermediate $q_k$ are integrated independently, so the contributing quantum paths need not be smooth.
For $H(p,q)=\tfrac{p^2}{2m}+V(q)$, each momentum integral is the following Gaussian.
{\small\begin{align}\begin{aligned}
    \int_{\R}\tfrac{dp}{2\pi\hbar}\exp\Big\{\tfrac{i\epsilon}{\hbar}\big[p\dot q-\tfrac{p^2}{2m}\big]\Big\} &= \Big(\tfrac{m}{2\pi i\hbar\epsilon}\Big)^{1/2}\exp\Big\{\tfrac{i\epsilon}{\hbar}\tfrac{m\dot q^2}{2}\Big\}.
\end{aligned}\end{align}}
Absorbing the product of normalization factors into $\mathcal{D}q$ yields the configuration-space path integral below.
{\small\begin{align}\begin{aligned}
    F(q_f,t_f;q_i,t_i) &= \int_{q(t_i)=q_i}^{q(t_f)=q_f}\mathcal{D}q\,e^{\tfrac{i}{\hbar}S[q]}, \qquad S[q] = \int_{t_i}^{t_f}dt\,L(q,\dot q), \qquad L(q,\dot q) = \tfrac12m\dot q^2-V(q).
\end{aligned}\end{align}}
Thus every path contributes with complex amplitude $e^{\tfrac{i}{\hbar}S[q]}$ fixed by its action, while the transition probability is the modulus squared of the sum of all these amplitudes.
This Gaussian reduction to configuration space requires a Hamiltonian quadratic in the momenta.
\end{mytheorem}


\begin{mytheorem}[Time-ordered green functions and path integrals]
%
We repeat the same argument above for transition amplitued between initial and final states with insertions of operators at intermediate times.
Assume that $t_1>t_2$ and consider the transition amplitude $\langle q_f, t_f | \hat{q}(t_1)\hat{q}(t_2) | q_i, t_i \rangle$.
Repeating the same steps we find
{\small\begin{align}\label{eq:first_transition}
    \langle q_f, t_f | \hat{q}(t_1)\hat{q}(t_2) | q_i, t_i \rangle = \int_{q(t_i)=q_i}^{q(t_f)=q_f} \!\!\!\!\!\!\!\mathcal{D}q(t)\,\, q(t_1)q(t_2) e^{\frac{i}{\hbar}S[q(t)]},
\end{align}}
On the other hand, if $t_1< t_2$, considering the transition $\langle q_f, t_f | \hat{q}(t_2)\hat{q}(t_1) | q_i, t_i \rangle$ and repeating the same steps we find
{\small\begin{align}\label{eq:second_transition}
    \langle q_f, t_f | \hat{q}(t_2)\hat{q}(t_1) | q_i, t_i \rangle = \int_{q(t_i)=q_i}^{q(t_f)=q_f} \!\!\!\!\!\!\!\mathcal{D}q(t)\,\, q(t_2) q(t_1)e^{\frac{i}{\hbar}S[q(t)]}\,.
\end{align}}
The right-hand sides of \eqref{eq:first_transition} and \eqref{eq:second_transition} are identical, since $q(t_1),\,q(t_2)$ are just numbers and commute.
We thus conclude that the path integral formulation of quantum mechanics naturally gives rise to time-ordered products of operators, that is
{\small\begin{align}
    \langle q_f, t_f | T\{ \hat{q}(t_1) \cdots \hat{q}(t_n) \} | q_i, t_i \rangle = \int_{q(t_i)=q_i}^{q(t_f)=q_f} \mathcal{D}q(t)\, q(t_1)\cdots q(t_n) e^{\frac{i}{\hbar}S[q(t)]}\,.
\end{align}}
More generally, repeating the same argument for any set of operators $O_j[\hat{q}(t_j)]$ functions of the fundamental fields we have
{\small\begin{align}
    \langle q_f, t_f | T\{ \hat{O}_1(t_1) \cdots \hat{O}_n(t_n) \} | q_i, t_i \rangle = \int_{q(t_i)=q_i}^{q(t_f)=q_f}\!\!\!\!\!\!\! \mathcal{D}q(t)\, O_1(t_1)\cdots O_n(t_n) e^{\frac{i}{\hbar}S[q(t)]}\,.
\end{align}}
%
\end{mytheorem}


\begin{mytheorem}[The partition function]
%
The time-ordered Green function of a quantum theory contains all the information about the theory,
{\small\begin{align}\label{eq:green_functions_vs_path_integral_1D}
    G_{q_i,t_i}^{q_f,t_f}(t_1,\ldots,t_n):=\langle q_f, t_f | T\{ \hat{q}(t_1) \cdots \hat{q}(t_n) \} | q_i, t_i \rangle = \int_{q(t_i)=q_i}^{q(t_f)=q_f} \mathcal{D}q(t)\, q(t_1)\cdots q(t_n) e^{\frac{i}{\hbar}S[q(t)]}\,.
\end{align}}
We would like to compress this information in a single object.
To fix ideas, consider the simpler case of an infinite sequence of numbers $a_0, a_1, \ldots, a_n, \ldots$.
We can encode all this information in a single function $f(x)$, the generating function of the sequence, defined as
{\small\begin{align}
    f(x) = \sum_{n=0}^{\infty} a_n x^n, \qquad \quad a_n = \frac{1}{n!} \left. \frac{d^n f}{dx^n} \right|_{x=0}.
\end{align}}
The generating function $f(x)$ contains all the information about the sequence, since we can recover the numbers $a_n$ by taking derivatives of $f(x)$ at $x=0$.
In the same way, we can encode all the information about the quantum theory in a single object, the \emph{partition function} or \emph{generating functional}.
Following the previous example we could define
{\small\begin{align}
    Z[J] = \sum_{n=0}^{\infty} \tfrac{i^n}{n!} \int dt_1 \cdots dt_n\, J(t_1)\cdots J(t_n) G(t_1,\ldots,t_n)\,.
\end{align}}
This expression is not immediately useful, since we would need to know all the Green functions to compute $Z[J]$.
However, we can use the definition of Green functions as transition amplitudes or as path integrals to rewrite the generating functional in convenient forms.
First, using 
{\small\begin{align}
    G(t_1,\ldots,t_n) = \langle q_f, t_f | T\{ \hat{q}(t_1) \cdots \hat{q}(t_n) \} | q_i, t_i \rangle_H
    =
    \langle q_f | T\{ e^{-\tfrac{i}{\hbar} \hat{H}(t_f-t_i)} \hat{q}(t_1) \cdots \hat{q}(t_n) \} | q_i \rangle_S,
\end{align}}
and since the $J(t)$ are just numbers, we can write
{\small\begin{align}
    Z[J] &= \sum_{n=0}^{\infty} \tfrac{i^n}{n!} \int dt_1 \cdots dt_n\, J(t_1)\cdots J(t_n) \langle q_f | T\{ e^{-\tfrac{i}{\hbar} \hat{H}(t_f-t_i)} \hat{q}(t_1) \cdots \hat{q}(t_n) \} | q_i \rangle_S
    \\
    &= \langle q_f | T\left\{ e^{-\tfrac{i}{\hbar} \hat{H}(t_f-t_i)+i\int dt\, J(t) \hat{q}(t)} \right\} | q_i \rangle_S.
\end{align}}
Similarly, using the second expression in \eqref{eq:green_functions_vs_path_integral_1D} we have
{\small\begin{align}\begin{aligned}
    Z[J] &= \sum_{n=0}^{\infty} \tfrac{i^n}{n!} \int dt_1 \cdots dt_n\, J(t_1)\cdots J(t_n) \int \mathcal{D}q(t)\, q(t_1)\cdots q(t_n) e^{\frac{i}{\hbar}S[q(t)]}
    \\
    &=\int \mathcal{D}q(t)\, e^{\frac{i}{\hbar}S[q(t)]+ i\int dt\, J(t) q(t)}.
\end{aligned}
\end{align}}
%
Note we are glossing over initial and final conditions on purpose.
This a subtle point which we adress in the next section.
\end{mytheorem}



%============================================================
\subsection{Path integral for harmonic oscillators \& the $\epsilon$-prescription}
%=========================================================

We now use the path integral to compute the propagator of a harmonic oscillator, and show how the $\epsilon$-prescription arises naturally in this context upon demanding natural boundary conditions and invertibility of the Gaussian operator in the action. We also show how the $\epsilon$-prescription is related to the choice of vacuum in the canonical quantization approach.

\begin{mytheorem}[Naive computation of the path integral \todoinline{finish}]
Let us start with the path integral for a free harmonic oscillator in the presence of a source $J(t)$.
Glossing over boundary conditions for the moment, we can integrate by parts the action to get
{\small\begin{align}
    S[q,j]=\int\! dt\, \tfrac12 m\dot{q}^2 - \tfrac12 m\omega^2 q^2 + J(t)q(t) =
    +\left[\tfrac12 m q\dot{q}\right]_{t_i}^{t_f}
    +\int\! dt\, \tfrac12 q(t)\Big(\underbrace{-m\tfrac{d^2}{dt^2}-m\omega^2}_{=:\Delta}\Big)q(t) + J(t)q(t).
\end{align}}
The operator $\Delta$ is a Gaussian operator with spectrum bounded below, eigenvalues $\lambda_p=m(p^2-\omega^2)$ and eigenfunctions $e^{ipt}$.
In turn the path integral formulation of the generating functional reads
{\small\begin{align}
    Z[J] = \int \mathcal{D}q(t)\, e^{\frac{i}{\hbar}S[q,J]} 
    = e^{\frac{i}{\hbar}\left[\tfrac12 m q\dot{q}\right]_{t_i}^{t_f}}
     \int \mathcal{D}q(t)\, \,  e^{\frac{i}{\hbar}  \int dt\, \tfrac12 q(t)\Delta q(t) + J(t)q(t)}.
\end{align}}
The boundary term drops out of the path integral since it only depends on the fixed boundary conditions.
Moreover, assuming for the moment that the operator $\Delta$ is invertible, we can write $q:=\tilde{q}-\Delta^{-1}J$ and complete the square in the exponent to get
{\small\begin{align}
      \int \mathcal{D}q\, \,  e^{i \left(\tfrac12 q\cdot \Delta\cdot q + J\cdot q\right)} = \int \mathcal{D}\tilde{q}\, \,  e^{i \tfrac 12 \tilde{q}\Delta\tilde{q} - \frac{i}{2} J\cdot \Delta^{-1}\cdot J}
      = 
      e^{-\frac{i}{2} J\cdot \Delta^{-1}\cdot J}
      \int \mathcal{D}\tilde{q}\, \,  e^{\frac{i}{\hbar} \tfrac12 \tilde{q}\Delta\tilde{q}}.
\end{align}}
Note we are free to take the linear term out of the integral and change the measure $Dq\to D\tilde{q}$ since $J$ is not integrated over and $\tilde{q}$ is just a shift of the integration variable.
The remaining integral is a Gaussian integral.
We can compute it explictly
{\small\begin{align}
    \int \mathcal{D}\tilde{q}\, \,  e^{\frac{i}{\hbar} \tfrac12 \tilde{q}\Delta\tilde{q}} = \mathcal{N}\det \Delta^{-1/2}, \quad \mathcal{N}\simeq \lim_{N\to\infty} \left(2\pi i\right)^{N/2}.
\end{align}}
The normalization constant $\mathcal{N}$ and the functional determinant $\det \Delta$ are ill-defined as the number $N$ of time slices go to infinity, resulting in infinitely many Gaussian factors $\prod_{j=1}^Ndq(t_j)$.
However, they are independent of the source $J$ and thus cancel out when computing correlation functions, which are ratios of generating functionals.
In fact, it is customary to absorb the factor $\mathcal{N}$ in the definition of the functioal determinant, since this will cancel anyway.

What is the operator inverse $\Delta^{-1}$?
Arguing formally, from the requirement it is indeed the inverse of $\Delta$ we have
{\small\begin{align}
    \Delta \Delta^{-1} = \delta_D^{(1)}(t-t') \quad \Rightarrow\quad \Delta^{-1}(t-t') = \int \tfrac{dp}{2\pi} \tfrac{e^{ip(t-t')}}{m(p^2-\omega^2)}.
\end{align}}
However, the expression is ill-defined since the integrand has a badly divergent pole at $p=\pm\omega$.
Of course we can slightly deform the contour of integration and avoid the pole with an infinitesimal complex shift $\pm i\epsilon$.
However, there are several inequivalent choices, and we want to understand what is their physical meaning.

We can already see this issue is related to the boundary conditions imposed on the path integral.
Indeed, the operator $\Delta$ is invertible if it has no zero eigenvalues.
The eigenfunctions of $\Delta$ are $e^{ipt}$ with eigenvalues $\lambda_p=m(p^2-\omega^2)$, and the zero eigenvalue corresponds to the eigenfunction $e^{i\omega t}$.
This is a solution of the classical equations of motion, and thus corresponds to a classical trajectory.
The boundary conditions are oscillations in the far past and far future, or equivalently PBC-like boundary conditions in  finite interval.
In order to make $\Delta$ invertible we should then impose boundary conditions that rule out these functions from the domain of $\Delta$.


\end{mytheorem}








%--------------------------------------------------------
%=========================================================
\section{Generating functionals}
%=========================================================
%------------------------------------------------------------


\begin{mytheorem}[Quantum theory is to classical theory what Fourier transform is to Legendre transform.\notetag{Polish}]
In formulas
    \begin{equation}
    \text{Quantum theory} \,\,:\,\, \text{classical theory} = \text{Fourier transform}\,\, :\,\, \text{Legendre transform}
\end{equation}
The analogy is made precise in the path integral formulation of QFT via the generating functional $W$ of a field theory
\begin{align}
    e^{iW[J]} &= \int \mathcal{D}\phi\, e^{iS[\phi]} e^{-i\int d^4x\, J(x)\phi(x)}.
\end{align}
In this sense $W[J]$ is the infinite-dimensional Fourier transform of the functional $\phi\mapsto e^{iS[\phi]}$ and $J(x)$ is the conjugate variable to the field $\phi(x)$.
The generating functional is the generator of all connected diagrams, and its knowledge is thus equivalent to the knowledge of the full quantum theory.

The \emph{stationary phase approximation} consists in approximating the path integral by the configuration that contributes the most to the integral, i.e. the one extremizing the action and thus reducing the oscillations of the integrand,
\begin{align}
    &\int \mathcal{D}\phi\, e^{iS[\phi]-\int d^4x\, J(x)\phi(x)} \sim e^{iS[\phi_{cl}]-\int d^4x\, J(x)\phi_{cl}(x)}\quad \text{for $\,\phi_{cl}\,$ solving}\quad \left.\frac{\delta S}{\delta \phi(x)}\right|_{\phi_{cl}} = J(x)\\
    &\quad \Rightarrow \quad W[J] \sim S[\phi_{cl}] - \int d^4x\, J(x)\phi_{cl}(x)\,.
\end{align} 
By definition this approximates $W[J]$ with the Legendre transform of the classical action.

Finally, recalling that the \emph{effective action} $\Gamma[\phi]$ is defined as the Legendre transform of $W[J]$
\begin{align}
    \Gamma[\phi] = W[J] - \int d^4x\, J(x)\phi(x) \quad \text{for $J$ solving}\quad \frac{\delta W}{\delta J(x)} = \phi(x) \quad \xrightarrow[]{\text{stationary ph.}}\quad \Gamma[\phi] \sim S[\phi].
\end{align}
we see that in stationary phase approximation the effective action reduces to the classical action.

To make the analogy clear, consider the 1-dimensional Fourier transform of a function $f(y) = e^{iS(y)}$, where $S(y), g(z)\in\C$ so that there is no loss of generality in the exponentiation,
\begin{align}
    e^{ig(z)}=\int dy\, e^{iS(y)} e^{-izy}.
\end{align}
Then, in stationary phase approximation, we approximate the Fourier transform with the Legendre transform
\begin{align}
    &\int dy\, e^{iS(y)-izy} \sim e^{iS(y_{cl}) - izy_{cl}}\quad \text{for $\,y_{cl}\,$ solving}\quad \left.\frac{dS}{dy}\right|_{y_{cl}} = z \quad \Rightarrow \quad g(z) \sim S(y_{cl}) - zy_{cl}.
\end{align}
%
\end{mytheorem}
\end{document}
