% !TeX root = ../QFT_main.tex
\documentclass[../QFT_main.tex]{subfiles}
\begin{document}
%=========================================================
%=========================================================
\chapter{Renormalization}\label{ch:renormalization}
%========================================================
%=========================================================



%--------------------------------------------------------
%=========================================================
\section{Generalities on renormalization}
%=========================================================
%------------------------------------------------------------


\begin{mytheorem}[What is renormalization? (cf. Matheus 2.1)]
Renormalization is not merely `absorbing divergences into redefinitions of the parameters of the theory', in fact renormalization is not even about divergences at all.
Renormalization is the process of relating the parameters of a theory to physical observables, and it is necessary in any quantum field theory, even in those that are finite and thus do not require any regularization procedure.
Specifically when writing down a Lagrangian for a quantum field theory, parameters such as masses and coupling constants are not the physical ones, but rather `bare' parameters that are not directly measurable.
Renormalization is the process of relating these bare parameters to the physical ones, which are the ones that can be measured in experiments.
In this sense, renormalization is a crucial step in making predictions from a quantum field theory and connecting it to the real world.
\end{mytheorem}




\begin{mytheorem}[Power counting and renormalizability]
To be discussed.
\end{mytheorem}



%--------------------------------------------------------
%=========================================================
\section{Renormalization and symmetries}
%=========================================================
%---------------------------------------------------------



%--------------------------------------------------------
%=========================================================
\section{Wilsonian renormalization and the renormalization group}
%=========================================================
%------------------------------------------------------------



%--------------------------------------------------------
%=========================================================
\section{Renormalization of QED}
%=========================================================
%---------------------------------------------------------


%--------------------------------------------------------
%=========================================================
\section{Renormalization of Non-Abelian gauge theories}
%=========================================================
%---------------------------------------------------------


\end{document}
